Energy Clustering

### 2017-10-08 energy_clustering/draft/energy_clustering.tex
more comments

% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
% tradition that most referees like to see. In this case I also included 
% some important information about our work, and it helps to make a transition 
% between introduction and the remaining of the paper. 
% I would leave it, but if you prefer to remove it, let me know and I 
% will adapt the paper.

% @gui: my experience is that most reviewers absolutely do *NOT* want it, rather, we skip it, and journals often remove it prior to publication.

@@

% @jovo: check this paragraph. I included it to answer your question
% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
% the term K(x,x) will be hidden inside the mean of the cluster to which
% x is currently assigned to. This also clarifies the connection to 
% the standard formulation of kernel k-means.
Obviously, $\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
For each $D$ we generate $100$ Monte Carlo runs, 
sampling $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the accuracy \eqref{eq:accuracy}. The algorithms
$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
both use the standard
kernel \eqref{eq:standard_metric}, as well as the spectral
$\mathcal{E}^S$-clustering method\footnote{We
compared our implementation of spectral clustering provided in
Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
by \emph{scikit-learn} library \cite{scikit-learn}, 
which is based on \cite{Malik}. In the few
cases considered the algorithms had similar performance.
% @gui: dude, this is a crazy sentence to write. use sklearn, or have a *very* good reason not to.  
% currently, you utterly fail to justify the difference.
}, 
and we initialize the algorithms
with the initialization of $k$-means++ \cite{Vassilvitskii}.
% @gui: i thought you were goign to initialize E-H & E-L with E-S?
For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
while GMM is initialized with the output of $k$-means. Every algorithm
is ran $5$ times, returning the answer with best objective function value.
We plot the average accuracy versus $D$, with error bars indicating
standard error.
The results are shown in Fig.~\ref{fig:gauss}a. We can see that
$\mathcal{E}^H$- and $\mathcal{E}^S$-clustering have practically
the same performance, which is 
much higher than $\mathcal{E}^L$-clustering (kernel
$k$-means). Moreover, $\mathcal{E}^H$-clustering outperforms 
$k$-means and GMM, where the improvement is noticeable specially in 
high dimensions.
Note that in this setting $k$-means and GMM are consistent estimators,
however, energy clustering degrades much less as dimension increases.
% @gui: k-means is only consistent if the prob of each class is equal.  but you did not actually define this parameter, because you sampled 100 from each.  so, you gotta define the parameter at least if you want to say anythng about consistency.  also, what do you mean k-means is a consistent estimator? estimator of what? clarify.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of energy clustering algorithms to $k$-means and GMM in 
high dimensional Gaussian settings. We plot the mean accuracy versus
the number of dimensions, with error bars indicating standard 
error from $100$ Monte
Carlo runs.
(a) Data distributed as in \eqref{eq:gauss1}, with Bayes
accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
(b) Data distributed
as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
over the range $D \in [10, 700]$.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
% @jovo: didn't understand this
}
\end{figure}

We now allow the diagonal covariance of the second Gaussian distribution to
have different values by choosing
$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
we consider the fixed setting
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637),
\end{split}
\end{equation}
as we vary $D$. Note that there is nothing special about the numbers in
$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
as well.
% @gui: if there is nothing special, why not just use 2*ones?  this seems suspicious.  also, it is the result of sampling 1 time, not 10 times.
The Bayes error is kept fixed yielding
an optimal accuracy $\approx 0.95$.
% @gui: bayes error changes for each random sample, unless you are integrating over all possible samples?  it is not fixed.
In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
$k$-means, and GMM.
Note that GMM performs better in low dimensions, 
since it is a consistent estimator.
However, $\mathcal{E}^H$ and $\mathcal{E}^S$ degrade much less as dimensions
increase, performing better than $k$-means and GMM in high dimensions.
% @gui: you gotta explain why GMM doesn't utterly fail when D>n!

>>>>

% @gui: my experience is that most reviewers absolutely do *NOT* want a ToC, rather, we skip it, and journals often remove it prior to publication.

@@

% @jovo: check this paragraph. I included it to answer your question
% about having K(x,x) and \ell=1,...,k without exluding \ell=j. Essentially,
% the term K(x,x) will be hidden inside the mean of the cluster to which
% x is currently assigned to. This also clarifies the connection to 
% the standard formulation of kernel k-means.
$\mathcal{E}^{L}$-clustering is precisely
kernel $k$-means algorithm \cite{Dhillon2,Dhillon} 
but written more concisely and with the kernel induced by energy statistics. Indeed, recalling that
$\kk(x,y)=\langle \varphi(x), \varphi(y) \rangle$ where $\varphi: \mathcal{X}
\to \mathcal{H}_\kk$ is the feature map, we have from \eqref{eq:Jell} that
\begin{equation}
%\begin{split}
J^{(\ell)}(x_i) 
%&= \dfrac{1}{n_\ell^2} \sum_{x, y \in \C_\ell} 
%\langle \varphi(x), \varphi(y) \rangle - \dfrac{2}{n_\ell}
%\sum_{y\in \C_\ell} \langle \varphi(x_i), \varphi(y)\rangle \\ 
= 
\langle \varphi(\mu_\ell), \varphi(\mu_\ell) \rangle
-2 \langle \varphi(x_i), \varphi(\mu_\ell) \rangle 
= 
\| \varphi(x_i) - \varphi(\mu_\ell) \|^2 - \| \varphi(x_i) \|^2,
%\end{split}
\end{equation}
where $\mu_\ell = \tfrac{1}{n_\ell} \sum_{x\in \C_\ell}x$ 
is the mean of cluster $\C_\ell$. 
Therefore, $\min_\ell J^{(\ell)}(x_i) = 
\min_\ell \| \varphi(x_i) - \varphi(\mu_\ell)\|^2$, i.e. we are assigning
$x_i$ to the cluster with closest center (in feature space),
which is the familiar Lloyd's
heuristic approach that kernel $k$-means is based upon.

@@

From the results of \cite{Telgarsky}, summarized in the end of the previous
section, we expect the improvement of Hartigan's 
over Lloyd's method to be more accentuated in high dimensions.
Thus, we analyze
how the algorithms degrade as the number of dimensions increase while
keeping the number of points in each cluster fixed.
Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, yielding an optimal 
accuracy of $\approx 0.86$.
For each $D$ we generate $100$ Monte Carlo runs, 
sampling $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the accuracy \eqref{eq:accuracy}. The algorithms
$\mathcal{E}^H$- and $\mathcal{E}^L$-clustering
both use the standard
kernel \eqref{eq:standard_metric}, as well as the spectral
$\mathcal{E}^S$-clustering method\footnote{We
compared our implementation of spectral clustering provided in
Algorithm~\ref{spectralalgo} with the spectral clustering routine provided
by \emph{scikit-learn} library \cite{scikit-learn}, 
which is based on \cite{Malik}. In the few
cases considered the algorithms had similar performance.
% @gui: dude, this is a crazy sentence to write. use sklearn, or have a *very* good reason not to.  
% currently, you utterly fail to justify the difference.
}, 
and we initialize the algorithms
with the initialization of $k$-means++ \cite{Vassilvitskii}.
% @gui: i thought you were goign to initialize E-H & E-L with E-S?
% @gui: you gotta clearly define E-S, with a pseudocode block, just like the others.  in particular, which matrix are you doing an eigendecomposition on, and how you are choosing the number of dimensions to embed into.
For $k$-means and GMM we use the implementation of \emph{scikit-learn} library
\cite{scikit-learn}, where $k$-means is also initialized with $k$-means++,
while GMM is initialized with the output of $k$-means. Every algorithm
is ran $5$ times, returning the answer with best objective function value.
We plot the average accuracy versus $D$, with error bars indicating
standard error.
The results are shown in Fig.~\ref{fig:gauss}a. We can see that
$\mathcal{E}^H$- and $\mathcal{E}^S$-clustering have practically
the same performance, which is 
much higher than $\mathcal{E}^L$-clustering (kernel
$k$-means). Moreover, $\mathcal{E}^H$-clustering outperforms 
$k$-means and GMM, where the improvement is noticeable specially in 
high dimensions.
Note that in this setting $k$-means and GMM are consistent estimators,
however, energy clustering degrades much less as dimension increases.
% @gui: k-means is only consistent if the prob of each class is equal.  but you did not actually define this parameter, because you sampled 100 from each.  so, you gotta define the parameter at least if you want to say anythng about consistency.  also, what do you mean k-means is a consistent estimator? estimator of what? clarify.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_mean.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{normal_highdim_cov.pdf}\\[-1.0em] (b)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of energy clustering algorithms to $k$-means and GMM in 
high dimensional Gaussian settings. We plot the mean accuracy versus
the number of dimensions, with error bars indicating standard 
error from $100$ Monte
Carlo runs.
(a) Data distributed as in \eqref{eq:gauss1}, with Bayes
accuracy $\approx 0.86$, over the range $D \in [10,200]$. 
(b) Data distributed
as in \eqref{eq:gauss2}, with Bayes accuracy $\approx 0.95$,
over the range $D \in [10, 700]$.
% @gui: also helpful here would be pairs plots, say, of the first 5 dimensions, showing the performance of our stuff (color points by Alg 3, maybe symbols denote "truth".  i have no intuition how difficult these are.
% @jovo: didn't understand this
% @gui: https://www.google.com/search?q=pairs+plots&rlz=1C5CHFA_enUS731US731&source=lnms&tbm=isch&sa=X&ved=0ahUKEwjwo9u5w-DWAhXMKyYKHYkhAKMQ_AUICigB&biw=1440&bih=776
% to show how well separated the classes are.
}
\end{figure}

% @gui: for eq 59, E-H & E-S yield the same performance.  but for eq 63, E-H is much better than E-S.  do you have any idea why that is?
We now allow the diagonal covariance of the second Gaussian distribution to
have different values by choosing
$(\Sigma_2)_{ii} \stackrel{iid}{\sim} \textnormal{Uni}(1,5)$ for
$i=1,\dotsc,10$, and $(\Sigma_2)_{ii} = 1$ for $i=11,\dotsc,D$. For instance,
we consider the fixed setting
\begin{equation}
\label{eq:gauss2}
\begin{split}
x & \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad
\Sigma_2 = \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,
1.969, 4.035,   4.237,  2.813,  3.637),
\end{split}
\end{equation}
as we vary $D$. Note that there is nothing special about the numbers in
$\widetilde{\Sigma}_{10}$, it is simply a random choice of sampling
$\textnormal{Uni}(1,5)$ ten times, and one could pick any other choices
as well.
% @gui: if there is nothing special, why not just use 2*ones?  this seems suspicious.  also, it is the result of sampling 1 time, not 10 times.
The Bayes error is kept fixed yielding
an optimal accuracy $\approx 0.95$.
% @gui: bayes error changes for each random sample, unless you are integrating over all possible samples?  it is not fixed.
In Fig.~\ref{fig:gauss}b we show the results of energy clustering,
$k$-means, and GMM.
Note that GMM performs better in low dimensions, 
since it is a consistent estimator.
However, $\mathcal{E}^H$ and $\mathcal{E}^S$ degrade much less as dimensions
increase, performing better than $k$-means and GMM in high dimensions.
% @gui: you gotta explain why GMM doesn't utterly fail when D>n!
